### 2020-06-29 latex/vnhcs.tex
Edited the preliminaries to use B_i a column and rephrased a lot of stuff.

\chapter{Development of Virtual Nonholonomic Constraints}
\section{Preliminaries on Analytical Mechanics}
In classical mechanics, a mechanical system is comprised of \(N\) point masses.
Each point mass represents the center of mass of a physical body, where the
position of each mass is described using three cartesian coordinates (one for
each spatial axis).  The system as a whole is represented by a vector in
\(\R^{3N}\), and the dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this method works for simple cases, it unfortunately proves impossible to
apply to a majority of mechanical systems since most of the forces are not known
explicitly.

Most mechanical systems are not solely a collection of point masses which can 
freely move about three-dimensional space; the masses are usually restricted
physically by some equations of constraint (EOC). 
Rather than representing the system by one vector in \(\R^{3N}\) and \(k\) EOC, 
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - k\) is the number of
\textit{degrees of freedom} (DOF) of the system \cite{greenwood_dynamics}.

Each generalized coordinate \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to write these as a \textit{configuration} vector 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system has the
following structure:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity}, and the combined vector
\((q,\dot{q})\) is called a \textit{state} of the system.

The field of analytical mechanics provides a computational method for finding
the EOM of a system in generalized coordinates. The two most common methods for
modelling robotic systems are \textit{Lagrangian} and \textit{Hamiltonian}
mechanics.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}:
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
In actuated mechanical systems, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^k\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i(q) \tau
\end{equation}
Here, \(B_i: \mathcal{Q} \rightarrow \R^{1\times k}\)
and the matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = [B_i(q)]
\]
is called the \textit{input matrix} for the system. 
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \(n - k\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
If this is the case, the Euler-Lagrange equations reduce to (\ref{eqn:el-eqns}),
\begin{equation}\label{eqn:el-eqns}
    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
\end{equation}
where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
\[
    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
    \left(\pdiff{D_{i,j}}{q_k}  +
     \pdiff{D_{i,k}}{q_j} -
     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
\]

@@

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n)\) and say that \(p\) is
the \textit{conjugate of momenta to \(q\)} Note that each \(p_i\) is a linear
function of \(\dot{q}\), so one can typically solve for \(\dot{q}(q,p)\) by
inverting the expressions from (\ref{eqn:p-i}). The combined vector \((q,p)\) is
called the \textit{phase} of the system.

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \cite{landau_mechanics}:
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations called \textit{Hamilton's equations}
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times k}\) is the input matrix and 
\(\tau \in \R^k\) is the vector of generalized input forces. A pair of
coordinates \((q,p)\) which satisfy Hamilton's equations are said to be 
\textit{canonical coordinates for} \(\mathcal{H}\). a change of coordinates
\((\hat{q}(q,p),\hat{p}(q,p))\) is said to be a \textit{canonical
transformation} if the EOM \((\dot{\hat{q}},\dot{\hat{p}})\) satisfy Hamilton's
equations for some other Hamiltonian \(\hat{H}\).

If the Kinetic energy is quadratic as in (\ref{eqn:lagrangian}), the conjugate of
momenta to \(q\) can be computed explicitly by
\[
    p = D(q)\dot{q}
\]
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p} = -\frac{1}{2} \left[p\tpose \pdiff{D^{-1}}{q_i}(q) p\right] 
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given an arbitrary Hamiltonian mechanical system (\ref{eqn:hamiltonian}) in
canonical coordinates \((q,p)\), it is not obvious how the input forces \(\tau\)
are affecting the conjugate of momenta \(p_i\). This is because \(\tau\) is
transformed by \(B(q)\), which may be quite complicated.

For this reason, we will define a new class of Hamiltonian systems with simple
input matrices, so that the effect of the input forces on the conjugate of
momenta is obvious. This class of systems will be the backbone for the rest of
the theory developed in this thesis.

\begin{defn}
    A Hamiltonian mechanical system (\ref{eqn:hamiltonian}) is said to be 
    \textit{simply actuated} if the input matrix \(B(q) \in \R^{n \times k}\)
    is always of the form
    \[
        B(q) = \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
\end{defn}

Under the following mild assumptions on the input matrix, we will show that
there exists a canonical transformation under which the system is simply
actuated.

\begin{assm}\label{assm:B-const}
    The input matrix \(B : \mathcal{Q} \rightarrow \R^{n\times k}\) is constant
    and full rank.
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a right-semi-orthogonal matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is a left-annihilator for \(B\). That is,
    \(B^\perp(B^\perp)\tpose = \Id{(n-k)}\) (right semi-orthogonality) and
    \(B^\perp B = \Zmat{(n-k) \times k}\) (left-annihilation).
\end{assm}

@@

\begin{assm}\label{assm:B-orthogonal}
    WLOG, the input matrix \(B\) is left semi-orthogonal. 
    That is, \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant full-rank matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{n \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\).
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau}\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator or \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal (i.e. it is full-rank and has inverse equal to its transpose):
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    \Rightarrow
    \mathbf{B}^{-1} = \mathbf{B}\tpose
\]

@@

In these new coordinates, the generalized input forces directly affect the
actuated conjugate of momenta. That is, \(\tau_i\) only affects \(p_{a,i}\), 
and all \(p_u\) are unaffected by the input forces. We prove this fact through
the following theorem.

\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the change of coordinates 
    \((Q = (q_u, q_a), P = (p_u,p_a))\) 
    is a canonical transformation of the
    original Hamiltonian system (\ref{eqn:hamiltonian}) into
    the simply actuated Hamiltonian system (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{\mathcal{H}}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2}
                \left[P\tpose \pdiff{\Minv}{Q_i}(Q) P \right] 
                - \nabla_Q V(Q) + 
                \begin{bmatrix}
                    \Zmat{(n-k)\times k} \\
                    \Id{k} \\
                \end{bmatrix} \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    For any constant matrix \(A\), the transformation
    \((Q = Aq, P = Ap)\) satisfies
    \(\pdiff{Q_i}{p_s} = \pdiff{P_i}{q_s} = 0\) for all 
    \(i,s \in \{1,\ldots,n\}\). Hence, the Poisson brackets between the
    new coordinates are zero:
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{s = 1}^n \pdiff{Q_i}{p_s}\pdiff{Q_j}{q_s} - 
        \pdiff{Q_i}{q_s}\pdiff{Q_j}{p_s} = 0 \\
        [P_i,P_j] &:= \sum\limits_{s=1}^n \pdiff{P_i}{p_s}\pdiff{P_j}{q_s} -
        \pdiff{P_i}{q_s}\pdiff{P_j}{p_s} = 0
    \end{align*}
    Since the matrix \(A = \mathbf{B}\) is invertible and orthogonal,
    \(A_i (A\tpose_j)\tpose = A_i (A^{-1}_j)\tpose = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between the new momenta and
    configurations are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{s=1}^n\pdiff{P_i}{p_s}\pdiff{Q_j}{q_s}
        - \pdiff{P_i}{q_s}\pdiff{Q_j}{p_s} \\
                  &= \sum\limits{s=1}^n A_{i,s}A_{j,s} - 0 \\
                  &= \sum\limits{s=1}^n A_{i,s}A\tpose_{s,j} \\
                  &= A_i (A\tpose_j)\tpose \\
                  &= \delta_{i,j}
    \end{align*}

@@

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \hat{B} = \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the new Hamiltonian system is simply actuated, as desired.
\end{proof}

>>>>

\chapter{Development of Virtual Nonholonomic Constraints}
\section{Preliminaries on Analytical Mechanics}
A mechanical system can be represented by \(N\) point masses, where each point
represents the center of mass of a physical body, along with \(r\)
\textit{equations of constraint} (EOC) which model the physical restrictions
between these masses.
The position of each point mass is described using three cartesian coordinates (one
for each spatial axis), so the system as a whole can be described by a vector in
\(\R^{3N}\) with \(r\) EOC. 
The dynamics of the system are computed by deriving the \(3N\)
\textit{equations of motion} (EOM) produced by Newton's second law \(F = m a\).
While this technique works for simple systems, it is impossible to
apply to a majority of mechanical systems since most of the forces 
on the system are not explicitly known.

Rather than modeling a mechanical system with point masses and constraints,
it is often feasible to represent the position of the system using \(n\)
independent scalar-valued variables \(q_1,\ldots,q_n\) called 
\textit{generalized coordinates}, where \(n = 3N - m\) is the number of
\textit{degrees of freedom} (DOF) of the system \cite{greenwood_dynamics}.

Each generalized coordinate \(q_i\) takes values in \(\Rt{T_i}\), where
\(T_i = \infty\) if \(q_i\) represents a length or \(T_i = 2\pi\) if \(q_i\)
represents an angle.
It is convention to collect the coordinates into a \textit{configuration} 
\(q = (q_1,\ldots,q_n) \in \mathcal{Q}\) 
where the \textit{configuration manifold} \(\mathcal{Q}\) of the system has the
following structure:
\[
    \mathcal{Q} = \Rt{T_1} \times \cdot \times \Rt{T_n}
\] 
The derivative \(\dot{q} = (\dot{q}_1,\ldots,\dot{q}_n)\) of a configuration
is called a \textit{generalized velocity}, and the combined vector
\((q,\dot{q})\) is called a \textit{state} of the system.

The field of analytical mechanics provides a computational method for finding
the EOM of a system in generalized coordinates. The two most common analytical
methods for modelling robotic systems are \textit{Lagrangian} and
\textit{Hamiltonian} mechanics.

@@

Lagrangian mechanics uses the kinetic energy \(T(q,\dot{q})\) and potential
energy \(P(q)\) of the system to define the Lagrangian 
\(\mathcal{L} : \mathcal{Q}\times\R^n \rightarrow \R\) defined by
(\ref{eqn:lagrangian-general}) \cite{greenwood_dynamics}.
\begin{equation}\label{eqn:lagrangian-general}
    \mathcal{L}(q,\dot{q}) = T(q,\dot{q}) - P(q)
\end{equation}
When the mechanical system is actuated, the EOM are described by \(n\) second-order
ordinary differential equations (ODEs) obtained from the \textit{Euler-Lagrange
equations} with \textit{generalized input forces} \(\tau \in \R^k\) 
(\ref{eqn:el-eqns-general}). 
\begin{equation}\label{eqn:el-eqns-general}
    \diff{}{t}\left\{ \pdiff{\mathcal{L}}{\dot{q}_i} \right\}
    - \pdiff{\mathcal{L}}{q_i} = B_i\tpose(q) \tau
\end{equation}
The vector \(B_i\tpose: \mathcal{Q} \rightarrow \R^{1\times k}\) describes how
the input forces shape the dynamics of \(q_i\).
The matrix  \(B: \mathcal{Q} \rightarrow \R^{n \times k}\) with
\[
    B(q) = [B_i\tpose(q)]
\]
is called the \textit{input matrix} for the system.
If \(k < n\), we say the system is \textit{underactuated} with degree of
underactuation \(n - k\).

Many actuated mechanical systems have quadratic kinetic energies, so that the
Lagrangian can be written explicitly as
\begin{equation}\label{eqn:lagrangian}
    \mathcal{L}(q,\dot{q}) = \frac{1}{2} \dot{q}\tpose D(q) \dot{q} - P(q)
\end{equation}
where the \textit{inertia matrix} \(D: \mathcal{Q} \rightarrow \R^{n\times n}\) 
is a symmetric, positive definite matrix for all \(q \in \mathcal{Q}\) and the
potential function \(P : \mathcal{Q} \rightarrow \R\) is smooth. 
%If this is the case, the Euler-Lagrange equations reduce to (\ref{eqn:el-eqns}),
%\begin{equation}\label{eqn:el-eqns}
%    D(q)\ddot{q} + C(q,\dot{q})\dot{q} + \nabla P(q) = B(q)\tau
%\end{equation}
%where the \textit{Coriolis matrix} \(C(q,\dot{q})\) is of the form
%\[
%    [C]_{i,j} = \frac{1}{2}\sum\limits_{k = 1}^n 
%    \left(\pdiff{D_{i,j}}{q_k}  +
%     \pdiff{D_{i,k}}{q_j} -
%     \pdiff{D_{k,j}}{q_i}\right)\dot{q}_k
%\]

@@

To do this, we first define the \textit{conjugate of momentum \(p_i\) to \(q_i\)} by
\begin{equation}\label{eqn:p-i}
    p_i(q,\dot{q}) = \pdiff{\mathcal{L}}{\dot{q}_i}(q,\dot{q})
\end{equation}
To ease notation, we will write \(p = (p_1, \ldots, p_n)\) and say that \(p\) is
the \textit{conjugate of momenta to \(q\)}. 
Note that each \(p_i\) is a linear function of \(\dot{q}\), so one can typically
solve for \(\dot{q}(q,p)\) by inverting the expressions from (\ref{eqn:p-i}).
The combined vector \((q,p)\) is called the \textit{phase} of the system.

Next we define the \textit{Hamiltonian} (\ref{eqn:hamiltonian-general}) 
as the Legendre transform of the Lagrangian \cite{landau_mechanics}.
\begin{equation}\label{eqn:hamiltonian-general}
    \mathcal{H}(q,p) = p\tpose \dot{q}(q,p) - \mathcal{L}(q,\dot{q}(q,p))
\end{equation}
The EOM in this framework can be shown to be the \(2n\)
first-order equations called \textit{Hamilton's equations}
\begin{equation}\label{eqn:hamiltons-eqns}
    \begin{cases}
        \dot{q} = \nabla_p\mathcal{H} \\
        \dot{p} = -\nabla_q\mathcal{H} + B(q)\tau \\
    \end{cases}
\end{equation}
where \(B(q) \in \R^{n\times k}\) is the same input matrix used by the
Lagrangian and \(\tau \in \R^k\) is the vector of generalized input forces.

If the kinetic energy is quadratic as in (\ref{eqn:lagrangian}), the conjugate of
momenta to \(q\) can be computed explicitly:
\[
    p = D(q)\dot{q}
\]
The resulting Hamiltonian system reduces to (\ref{eqn:hamiltonian}).
\begin{align}\label{eqn:hamiltonian}
    \mathcal{H}(q,p) &= \frac{1}{2} p\tpose D^{-1}(q) p + P(q) \\
                     &\begin{cases}
        \dot{q} = D^{-1}(q)p \\
        \dot{p} = -\frac{1}{2} \left[p\tpose \pdiff{D^{-1}}{q_i}(q) p\right] 
        - \nabla_q P(q) + B(q) \tau \\
    \end{cases} 
\end{align}

A pair of coordinates \((q,p)\) which satisfy Hamilton's equations 
under the Hamiltonian \(\mathcal{H}\) are
said to be \textit{canonical coordinates for} \(\mathcal{H}\).  A change of
coordinates \((q,p) \rightarrow (Q,P)\) is said to be a \textit{canonical
transformation} if \((Q,P)\) are canonical coordinates for
\(H(Q,P) = \mathcal{H}\left(q(Q,P), p(Q,P)\right)\).

\section{Simply Actuated Hamiltonian Systems}
% TODO: Describe the change of coordinates to get into q_u/q_a mode and show
% that the actuator directly affects pa but not pu. Use M for simply actuated
% inertia, D for normal coordinates
Given a Hamiltonian mechanical system (\ref{eqn:hamiltonian}), it is not obvious
how the input forces \(\tau\) affect the conjugate of momenta \(p_i\). 
This is because \(\tau\) is transformed by the input matrix \(B(q)\), which may
be quite complicated.

We will define a new class of Hamiltonian systems where the effect of the input
forces on the conjugate of momenta is made obvious. This class of systems will
form the backbone for the rest of the theory developed in this thesis.

\begin{defn}
    A pair of canonical coordinates \((q,p)\) for \(\mathcal{H}\) are said to be
    \textit{simply actuated coordinates} if, in these coordinates, the input
    matrix \(B(q) \in \R^{n \times k}\) is of the form
    \[
        B(q) = \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
\end{defn}
\begin{defn}
    A Hamiltonian system is said to be \textit{simply actuated} if there exists
    a canonical transformation into simply actuated coordinates for
    \(\mathcal{H}\).
\end{defn}

Under the following assumptions on the input matrix, we will show that the
Hamiltonian system (\ref{eqn:hamiltonian}) is simply actuated.

\begin{assm}\label{assm:B-const}
    The input matrix \(B(q) \equiv B \in \R^{n\times k}\) is constant,
    full rank, and \(k \leq n\).
\end{assm}
\begin{assm}\label{assm:B-perp}
    There exists a matrix 
    \(B^\perp \in \R^{(n-k)\times n}\)
    which is right semi-orthogonal 
    (\ie \(B^\perp(B^\perp)\tpose = \Id{(n-k)}\))
    and which is a left-annihilator for \(B\)
    (\ie \(B^\perp B = \Zmat{(n-k) \times k}\)).
\end{assm}

@@

\begin{assm}\label{assm:B-orthogonal}
    Assume without loss of generality that the input matrix \(B\) is left
    semi-orthogonal.
    That is, assume \(B\tpose B = \Id{k}\).
\end{assm}
\begin{proof}
Since \(B\) is a constant matrix, 
it has a singular-value decomposition 
\(B = U \Sigma V\tpose\) where \(U^{-1} = U\tpose \in \R^{n \times n}\), 
\(V^{-1} = V\tpose \in \R^{k \times k}\), and \(\Sigma \in \R^{n \times k}\) is
defined by
\[
    \Sigma = \begin{bmatrix}
        \sigma_1 & 0 & \cdots & 0 & 0 \\
        0 & \sigma_2 & \cdots & 0 & 0 \\
        \vdots & & \ddots &  & \vdots \\
        0 & 0 & \cdots & 0 & \sigma_k \\
          & - & \Zmat{(n-k)\times k} & -  &\\
    \end{bmatrix}
\]
where \(\sigma_i \neq 0\) because \(B\) is full-rank.
\textbf{SOURCE}
Defining \(T \in \R^{k \times k}\) by
\[
    T = \begin{bmatrix}
        \frac{1}{\sigma_1} & 0 & \cdots & 0 \\
        0 & \frac{1}{\sigma_2} & \cdots & 0 \\
    \vdots & & \ddots & \vdots \\
    0 & 0 & \cdots & \frac{1}{\sigma_k} \\
    \end{bmatrix}
\]
and assigning the input forces to \(\tau = V T \hat{\tau}\), we get a new input
matrix for \(\hat{\tau} \in \R^k\) given by \(\hat{B} = B V T = U \Sigma T\) 
which is still constant and full-rank. In particular, 
\(\hat{B}\tpose \hat{B} = T\tpose \Sigma\tpose \Sigma\tpose T = \Id{k}\).
\end{proof}

Let \(\mathbf{B} \in \R^{n\times n}\) be the following matrix:
\[
    \mathbf{B} = 
    \begin{bmatrix}
        B^\perp \\
        B\tpose \\
    \end{bmatrix}
\]
Since \(B^\perp\) is a left annihilator of \(B\) and both \(B^\perp\) and
\(B\tpose\) are right semi-orthogonal, it is easy to show that \(\mathbf{B}\) is
orthogonal:
\[
    \mathbf{B}\mathbf{B}\tpose = 
    \begin{bmatrix}
        B^\perp (B^\perp)\tpose & B^\perp B \\
        (B^\perp B)\tpose & B\tpose B
    \end{bmatrix} = \Id{n}
    \Rightarrow
    \mathbf{B}^{-1} = \mathbf{B}\tpose
\]

@@

The following theorem shows that these new coordinates are simply actuated
coordinates, so that \(\tau_i\) only affects \(p_{a,i}\)
and all \(p_u\) are unaffected by the input forces.

\begin{thm}
    Under Assumptions \ref{assm:B-const},\ref{assm:B-perp}, and
    \ref{assm:B-orthogonal}, the change of coordinates 
    \((Q = (q_u, q_a), P = (p_u,p_a))\) 
    is a canonical transformation of the
    original Hamiltonian system (\ref{eqn:hamiltonian}) into
    the simply actuated Hamiltonian system (\ref{eqn:simple-hamiltonian}),
    \begin{align}\label{eqn:simple-hamiltonian}
        \hat{H}(Q,P) &= 
        \frac{1}{2} P\tpose \Minv(Q) P + V(Q) \\
       &\begin{cases}
            \dot{Q} = \Minv(Q)P \\
            \dot{P} = -\frac{1}{2}
                \left[P\tpose \pdiff{\Minv}{Q_i}(Q) P \right] 
                - \nabla_Q V(Q) + 
                \begin{bmatrix}
                    \Zmat{(n-k)\times k} \\
                    \Id{k} \\
                \end{bmatrix} \tau
        \end{cases} \nonumber
    \end{align}
    where
    \begin{align*}
        \Minv(Q) &= \mathbf{B}D^{-1}(\mathbf{B}\tpose Q)\mathbf{B}\tpose \\
        V(Q) &= P(\mathbf{B}\tpose Q) \\
    \end{align*}
\end{thm}
\begin{proof}
    For any constant matrix \(A\), the transformation
    \((Q = Aq, P = Ap)\) satisfies
    \(\pdiff{Q_i}{p_m} = \pdiff{P_i}{q_m} = 0\) for all 
    \(i,m \in \{1,\ldots,n\}\). Hence, the Poisson brackets between the
    new coordinates are zero:
    \begin{align*}
        [Q_i,Q_j] &:= \sum\limits_{m = 1}^n \pdiff{Q_i}{p_m}\pdiff{Q_j}{q_m} - 
        \pdiff{Q_i}{q_m}\pdiff{Q_j}{p_m} = 0 \\
        [P_i,P_j] &:= \sum\limits_{m=1}^n \pdiff{P_i}{p_m}\pdiff{P_j}{q_m} -
        \pdiff{P_i}{q_m}\pdiff{P_j}{p_m} = 0
    \end{align*}
    Since the matrix \(A = \mathbf{B}\) is invertible and orthogonal,
    \((A_i)\tpose A\tpose_j = (A_i)\tpose (A^{-1})_j = \delta_{i,j}\). Using this
    fact we see that the Poisson brackets between \(P\) and \(Q\) are given by:
    \begin{align*}
        [P_i,Q_j] &= \sum\limits_{m=1}^n\pdiff{P_i}{p_m}\pdiff{Q_j}{q_m}
        - \pdiff{P_i}{q_m}\pdiff{Q_j}{p_m} \\
                  &= \sum\limits_{m=1}^n A_{i,m}A_{j,m} - 0 \\
                  &= \sum\limits_{m=1}^n A_{i,m}A\tpose_{m,j} \\
                  &= (A_i)\tpose A\tpose_j \\
                  &= \delta_{i,j}
    \end{align*}

@@

    Furthermore, since \(\dot{P} = \mathbf{B}\dot{p}\), the new input matrix is
    given by 
    \[
        \mathbf{B}B = \begin{bmatrix}
            B^\perp B \\
            B\tpose B \\
        \end{bmatrix} = 
        \begin{bmatrix}
            \Zmat{(n-k)\times k} \\
            \Id{k}
        \end{bmatrix}
    \]
    so the coordinates \((q_u,q_a,p_u,p_a)\) are simply actuated coordinates for 
    \(\hat{H}\) as desired.
\end{proof}
